\documentclass[12pt, letterpaper]{article}
\include{eu}

\begin{document}

\section*{\textsl{Einstein's Universe} Problem Set 1}

This problem set is not to be handed in for credit. But it is due
before \textbf{Tuesday September 21}, when some of these problems
will appear in part, near-verbatim, on Term Exam 1.

\begin{problem}
How many cubic millimeters are there in a cubic centimeter? How many
cubic centimeters are there in a liter? How many liters are there in a
cubic meter?
\end{problem}

\begin{problem}
What is the maximum amount of cash you can obtain by
successfully robbing an armored truck?  Assume that it is packed with
twenties; that is, estimate an answer by considering (very approximately) the volume of a
truck, and (harder to estimate) the volume of a 20-dollar bill.  State
your assumptions and explain your work, but please don't attempt an
experiment.  Be sure to explain exactly how you estimated the volume
of a 20-dollar bill.  \emph{Hint: think of a stack of bills to
  estimate the volume.  Feel free to \emph{check} any part of your
  answer on the internet, but make sure you actually make a justified,
  quantitative estimate independently.}

Given your answer, do you think that many of the armored trucks in Manhattan
are fully packed with 20-dollar bills?
\end{problem}

\begin{problem}
How many molecules of air are there in one cubic millimeter?
Approximately! You might want to look up the following things: How
many molecules are there in a mole? What is the volume of a mole at
STP? What is STP? Express your answer in scientific notation, and only
give your answer to two digits of accuracy.

Now: How many molecules of water are there in one cubic millimeter?
Approximately! Water is $\mathrm{H}_2\mathrm{O}$; use that to figure
out the mass of one mole of water. Then use the density of water to
figure out what fraction of a mole is in one cubic millimeter.
\end{problem}

\begin{problem}
What is the energy (in joules $\J$) of a photon with a wavelength of
$\lambda = 0.5\,\mum$? You will have to look up the relationship
between energy $E$ and wavelength $\lambda$; it involves the speed of
light $c$ and Planck's constant $h$. How many photons per unit time
(number per second) does a $2\,\W$ LED emit if you assume that all
that power goes into $0.5\,\mum$ photons? You might have to look up
the definition of a watt $\W$. Use scientific notation to express your
answer and don't forget your units.
\end{problem}

\begin{problem}
What are the base units of the expression $n\,m\,v^2$, where $n$
has units of number per volume, $m$ has units of mass, and $v$ has
units of speed? The base units are $\kg$ (kilograms), $\m$ (meters),
and $\s$ (seconds).

Is the answer you get the same base units as energy or force or
pressure or anything else you know? Why do you think we are asking
this question?
\end{problem}

\end{document}
